---
title: "ClimWin_Covariate"
author: "Kai Frazer"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

#### References

Bailey and van de Pol 2025. Advanced climwin. For example of spatial climwin

Bailey and van de Pol 2016. climwin. For intro to climwin

van de Pol et al. 2016. For a little more climwin background

#### AI Disclosure

Claude chat that first suggested climwin as a package to explore, as well as some introductory code for using it: <https://claude.ai/share/d8438437-ed0e-4b85-9ab0-8306300d8731>

Claude chat for debugging search4update, tbh it wasn't very helpful and tasks were extremely menial: <https://claude.ai/share/943806c6-af4c-4462-b5c5-4cc032ed38eb>

## Packages & Data

```{r}
#| label: libraries

library(pacman)
p_load(lme4, climwin, terra, dplyr, tidyverse, zoo, lubridate)
```

```{r}
#| label: data for model
# note: duplicate of methods from Frequentist_Null.qmd

matrices <- readRDS('/Users/kailafrazer/Library/CloudStorage/OneDrive-USNH/Crab/GoMBlueCrabs/ModelReadyData/modelready_matrices.rds')

# extract data i want
crab <- matrices[["crab_abun"]]
points <- matrices[["points"]]
pop <- matrices[["pop"]]

# function to translate datat to long form
make.long <- function(wide, flatornot, colname) {
  
  if (flatornot == "not") { # for timev varying data
    long <- as.data.frame(wide) %>% mutate(cell_id = rownames(wide)) %>%
      pivot_longer(-cell_id, names_to = "year", values_to = colname)}
  
  if (flatornot == "flat") { # for static data
    long <- as.data.frame(wide)
    long <- mutate(long, cell_id = rownames(long))}
  
  return(long)
}

# make everything long
crab_long <- make.long(crab, "not", "count")
points_long <- make.long(points, "flat", "")
pop_long <- make.long(pop, "flat", "")

# join data
data <- crab_long %>%
  left_join(points_long, by = "cell_id") %>%
  left_join(pop_long, by = "cell_id") %>% rename("pop" = "wide")
```

```{r}
#| label: climate data for climwin

## oisst code from Covariate_Construction.qmd

oisst_daily <- readRDS('/Users/kailafrazer/Library/CloudStorage/OneDrive-USNH/Crab/Data/oisst_1982_202603.rds') %>% 
  mutate(date_time = ymd_hms(time), date = date(date_time), 
         year = year(date), month = month(date), 
         yearmon = as.yearmon(date),
         season = factor(case_when(month %in% c(12,1,2) ~ "Winter",
                            month %in% c(3,4,5) ~ "Spring",
                            month %in% c(6,7,8) ~ "Summer",
                            month %in% c(9,10,11) ~ "Fall"))) %>%
  dplyr::select(longitude, latitude, date_time, date, year, yearmon, month, season, sst) %>%
  mutate(longitude = case_when(as.numeric(longitude) < 0 ~ longitude,
                               as.numeric(longitude) > 0 ~ as.numeric(longitude) - 360))

oisst_daily_nona <- na.omit(oisst_daily) # now the monthly data shouldn't have NAs

oisst_monthly_nona <- oisst_daily_nona %>%
  group_by(longitude, latitude, yearmon, year, season, month) %>%
  summarize(monthly_sst = mean(sst), monthly_min = min(sst), monthly_max = max(sst)) %>%
  mutate(longitude = case_when(as.numeric(longitude) < 0 ~ longitude,
                               as.numeric(longitude) > 0 ~ as.numeric(longitude) - 360))

oisst_month_points <- right_join(oisst_monthly_nona, 
                                   points_long, by = c("longitude", "latitude")) %>% 
  mutate(date = as.Date(yearmon)) %>% ungroup() %>% dplyr::select(date, monthly_sst, cell_id) # %>%
  # filter(year(date) != 2026)

climdat <- oisst_month_points %>% filter(year(date) < 2026)
```

```{r}
#| label: bio data for climwin

crab_month <- readRDS("ModelReadyData/crab[i,m,y].rds")
crab_month_long <- as.data.frame(crab_month) %>% mutate(lonlat = rownames(crab_month)) %>%
      pivot_longer(-lonlat, names_to = "yearmon", values_to = "abun") %>%
      separate(lonlat, into = c("longitude", "latitude"), sep="_") %>% 
  mutate(longitude = as.double(longitude), latitude = as.double(latitude)) %>%
  left_join(points_long) %>%
  mutate(date = as.Date(as.yearmon(yearmon, format="%Y.%b"))) %>% dplyr::select(date, abun, cell_id)

# make sure there's enough run-up time for oisst lag
min_date <- min(climdat$date) + years(4)
biodat <- crab_month_long %>% filter(date <= max_date, date >= min_date)

# make sure we're only looking at the window of presences and not long tails of absences at the beginning or end
only_abuns <- filter(biodat, abun > 0)
min_biodate <- min(only_abuns$date)
max_biodate <- max(only_abuns$date)
biodat <- biodat %>% filter(date <= max_biodate, date >= min_biodate, year(date) < 2026) %>% mutate(year = year(date), year_numeric = as.numeric(year), year_factor = factor(year), cell_id = as.numeric(cell_id))
```

## Null model

```{r}
#| label: reconstructing model from Frequentist_Covariate.qmd

# climwin can only take lm, glm, lmer, glmer

moddat <- biodat 

# basemod <- glm.nb(abun ~ year_numeric + (1|cell_id), data=moddat) # convergence errors in glmer.nb and cell_id doesn't help for glm.nb
basemod.revision <- glm.nb(abun ~ year_numeric, data=moddat) # 4119 and resulted in some funky climwin analysis
basemod <- glmer.nb(abun ~ (1|year), data=moddat) # 4105
# basemod <- glmer.nb(abun ~ (1|year_numeric), data=moddat) # 4105

# i think it makes sense to keep basemod over basemod.revision because basemod.revision has a higher AIC given the package frameworks available to work with climwin. basemod.revision also results in less seasonal patterning of the climwin but similar general takeaways, making it less clear to interpret but leading to similar things

# the main thing to learn from basemod.revision is that it does favor more recent years more highly in change in AIC, which basemod does less, so i think it's good to keep in mind that we shouldn't be searching over a super broad window (as in search6)

# ## or if climwin can actually take tmb, then...
# 
# moddat <- biodat %>% mutate(year = year(date)) %>% left_join(points_long, by = c("cell_id")) %>%
#   left_join(pop_long, by = c("cell_id")) %>% rename(pop = wide)
# 
# # or wait if it can accept glmmTMB...
# basemod_tmb <- glmmTMB(abun ~ (1|year), ziformula = ~1, family = nbinom2(), data = moddat) # dang, aic really goes up for the monthly data over the yearly null # mmm except climwin doesn't accept a zi formula
```

## Search for lag window

A 4-year lag window takes \~3hr to run

```{r}
#| label: search1 - first try at a search

search1 <- slidingwin(xvar = list(sst = climdat$monthly_sst),
                      cdate = climdat$date,
                      bdate = biodat$date,
                      baseline = basemod_tmb,
                      cinterval = "month",
                      range = c(48,0),
                      type = "relative",
                      stat = "mean", # c("mean", "min)
                      func = "lin", # c("lin", "quad")
                      spatial = list(biodat$cell_id, climdat$cell_id)) # takes an hour or two
saveRDS(search1, "/Users/kailafrazer/Library/CloudStorage/OneDrive-USNH/Crab/slidingwin_search1_20260724.rds")

View(search1[[1]]$Dataset)
search1[[1]]$BestModel
search1[[1]]$BestModel@optinfo$conv$lme4$messages # did have failed to converge messages :(

out1 <- search1[[1]]$Dataset

# plots from climwin vignette
plotdelta(dataset = out1) # WOW there is seasonality
plotweights(dataset = out1) # very small number of models in the 95% confidence set
plotbetas(dataset = out1) # clear increase in effect a year out
plotwin(dataset = out1) # not quite sure what to make of these
```

```{r}
#| label: search2 - now with scaled values for temp in hopes that that'll throw fewer errors

search2 <- slidingwin(xvar = list(sst = as.numeric(scale(climdat$monthly_sst))),
                      cdate = climdat$date,
                      bdate = biodat$date,
                      baseline = basemod,
                      cinterval = "month",
                      range = c(48,0),
                      type = "relative",
                      stat = "mean", # c("mean", "min)
                      func = "lin", # c("lin", "quad")
                      spatial = list(biodat$cell_id, climdat$cell_id))

saveRDS(search2, "/Users/kailafrazer/Library/CloudStorage/OneDrive-USNH/Crab/slidingwin_search2_20260724.rds")

View(search2[[1]]$Dataset)
search2[[1]]$BestModel
search2[[1]]$BestModel@optinfo$conv$lme4$messages # did have failed to converge messages :(

out2 <- search2[[1]]$Dataset

# plots from climwin vignette
plotdelta(dataset = out2) # WOW there is seasonality
plotweights(dataset = out2) # very small number of models in the 95% confidence set
plotbetas(dataset = out2) # clear increase in effect a year out
plotwin(dataset = out2) # not quite sure what to make of these
```

```{r}
#| label: search3 - now with absolute values from jan 7

search3 <- slidingwin(xvar = list(sst = as.numeric(scale(climdat$monthly_sst))),
                      cdate = climdat$date,
                      bdate = biodat$date,
                      baseline = basemod,
                      cinterval = "month",
                      range = c(48,0),
                      type = "absolute",
                      refday = c(7,1),
                      stat = "mean", # c("mean", "min)
                      func = "lin", # c("lin", "quad")
                      spatial = list(biodat$cell_id, climdat$cell_id))

saveRDS(search3, "/Users/kailafrazer/Library/CloudStorage/OneDrive-USNH/Crab/slidingwin_search3_20260724.rds")

View(search3[[1]]$Dataset)
search3[[1]]$BestModel
search3[[1]]$BestModel@optinfo$conv$lme4$messages # no failure to converge messages, that's good

out3 <- search3[[1]]$Dataset

# plots from climwin vignette
plotdelta(dataset = out3) # WOW there is seasonality
plotweights(dataset = out3) # very small number of models in the 95% confidence set
plotbetas(dataset = out3) # clear increase in effect a year out
plotwin(dataset = out3) # not quite sure what to make of these
```

```{r}
#| label: search4update - now with absolute values from jul 1

# update means the model has been updated from the one i thought was best (just the random effect) to the one i realized i could do and was better (just year covariate)

# when i first ran this with all the new data from 2026, i was hitting a lot of window errors, i think because july 2026 wasn't available, and that's the refday
search4update <- slidingwin(xvar = list(sst = as.numeric(scale(climdat$monthly_sst))),
                      cdate = climdat$date,
                      bdate = biodat$date,
                      baseline = basemod,
                      cinterval = "month",
                      range = c(48,0),
                      type = "absolute",
                      refday = c(01,07),
                      stat = "mean", # c("mean", "min)
                      func = "lin", # c("lin", "quad")
                      spatial = list(biodat$cell_id, climdat$cell_id))

saveRDS(search4update, "/Users/kailafrazer/Library/CloudStorage/OneDrive-USNH/Crab/slidingwin_search4update_20260724.rds")

View(search4update[[1]]$Dataset)
search4update[[1]]$BestModel
search4update[[1]]$BestModel@optinfo$conv$lme4$messages # no convergence messages! yay

out4update <- search4update[[1]]$Dataset

# plots from climwin vignette
plotdelta(dataset = out4update) # WOW there is seasonality - okay, so the plot with just the year covariate is way less easy to distinguish patterns from. i think this is because the year covariate is already soaking up so much of the variation
plotweights(dataset = out4update) # very small number of models in the 95% confidence set
plotbetas(dataset = out4update) # clear increase in effect a year out
plotwin(dataset = out4update) # not quite sure what to make of these
```

```{r}
#| label: search5 - now looking at min effects instead of mean effects

search5 <- slidingwin(xvar = list(sst = as.numeric(scale(climdat$monthly_sst))),
                      cdate = climdat$date,
                      bdate = biodat$date,
                      baseline = basemod,
                      cinterval = "month",
                      range = c(48,0),
                      type = "absolute",
                      refday = c(01,07),
                      stat = "min",
                      func = "lin", # c("lin", "quad")
                      spatial = list(biodat$cell_id, climdat$cell_id))

saveRDS(search5, "/Users/kailafrazer/Library/CloudStorage/OneDrive-USNH/Crab/slidingwin_search5_20260724.rds")

View(search5[[1]]$Dataset)
search5[[1]]$BestModel
search5[[1]]$BestModel@optinfo$conv$lme4$messages

out5 <- search5[[1]]$Dataset

# plots from climwin vignette
plotdelta(dataset = out5)
plotweights(dataset = out5)
plotbetas(dataset = out5)
plotwin(dataset = out5)
```

```{r}
#| label: search6 - 10 year lag window

# adjust data so we have 10 years of climdat sans biodat
biodat10 <- filter(biodat, date %in% unique(oisst_month_points$date)[121:533]) %>% na.omit()

# creating null model for this biodat
moddat10 <- biodat10 %>% mutate(year = year(date))
basemod10 <- glmer.nb(abun ~ (1|year), data=moddat10) # AIC of 1722 when run on the yearly aggregated data (data), AIC of 4144

search6 <- slidingwin(xvar = list(sst = as.numeric(scale(climdat$monthly_sst))),
                      cdate = climdat$date,
                      bdate = biodat10$date,
                      baseline = basemod10,
                      cinterval = "month",
                      range = c(120,0),
                      type = "absolute",
                      refday = c(01,07),
                      stat = "min",
                      func = "lin", # c("lin", "quad")
                      spatial = list(biodat10$cell_id, climdat$cell_id)) # started 9:28AM

saveRDS(search6, "/Users/kailafrazer/Library/CloudStorage/OneDrive-USNH/Crab/slidingwin_search6_20260724.rds")

View(search6[[1]]$Dataset)
search6[[1]]$BestModel
search6[[1]]$BestModel@optinfo$conv$lme4$messages

out6 <- search6[[1]]$Dataset

# plots from climwin vignette
plotdelta(dataset = out6)
plotweights(dataset = out6)
plotbetas(dataset = out6)
plotwin(dataset = out6)
```

## Randomize to test the stat strength of the output

5 repeats for a 4-year lag window took about 24hrs with waking my computer up every couple hours and having it die overnight

```{r}
#| label: search4 -  randwin

# rerun the analysis with dates randomized to create a dataset of random ideal windows
rand4 <- randwin(repeats = 5, # 1 run was looking to take as long as a normal run of the climwin, which was about 1.5 hours, so i'm keeping to the minimum ex. used in the vignette
                    xvar = list(sst = as.numeric(scale(climdat$monthly_sst))),
                    cdate = climdat$date,
                    bdate = biodat$date,
                    baseline = basemod,
                    cinterval = "month",
                    range = c(48,0),
                    type = "absolute",
                    refday = c(01,07),
                    stat = "mean",
                    func = "lin",
                    spatial = list(biodat$cell_id, climdat$cell_id))

saveRDS(rand4, "/Users/kailafrazer/Library/CloudStorage/OneDrive-USNH/Crab/slidingwin_rand4_20260725.rds")

## get the p-value of our results
out4 <- search4[[1]]$Dataset # og output
out4rand = rand4[[1]]

pvalue(dataset = out4, datasetrand = out4rand, 
       metric = "C", # for small randomization samples (in the vignette they used "C" for 5 repeats of randwin)
       sample.size = 44) # 43 years used in analysis # 8.263909e-06
```
