---
title: "Frequentist_Covariate"
author: "Kai Frazer"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

#### References:

Zuur et al. 2009 for workflow to build a model (from the best null to the best single fixed effect to the best random effects to the best fixed structure)

Bolker 2020 for details to add random effects to glmmTMB model

#### AI Disclosure:

Claude chat helping me with little errors writing response curve <https://claude.ai/share/68065871-5bbc-4b06-878d-274b0e23e482>

## Packages & Data:

```{r}
#| label: libraries

library(pacman)
p_load(glmmTMB, ggplot2, dplyr, patchwork)
```

```{r}
#| label: data
# note: duplicate of methods from Frequentist_Null.qmd

matrices <- readRDS('/Users/kailafrazer/Library/CloudStorage/OneDrive-USNH/Crab/GoMBlueCrabs/ModelReadyData/modelready_matrices.rds')

# extract data i want
crab <- matrices[["crab_abun"]]
points <- matrices[["points"]]
depth <- matrices[["depth"]]
winter_mean <- matrices[["winter_mean"]]
summer_mean <- matrices[["summer_mean"]]
winter_min <- matrices[["winter_min"]]
summer_max <- matrices[["summer_max"]]
fall_flags <- matrices[["fall_flags"]]
winter_flags <- matrices[["winter_flags"]]
spring_flags <- matrices[["spring_flags"]]
yearly_flags <- matrices[["yearly_flags"]]
ddd <- matrices[["ddd"]]
rdd <- matrices[["rdd"]]
sdd <- matrices[["sdd"]]
pop <- matrices[["pop"]]

# function to translate datat to long form
make.long <- function(wide, flatornot, colname) {
  
  if (flatornot == "not") { # for timev varying data
    long <- as.data.frame(wide) %>% mutate(cell_id = rownames(wide)) %>%
      pivot_longer(-cell_id, names_to = "year", values_to = colname)}
  
  if (flatornot == "flat") { # for static data
    long <- as.data.frame(wide)
    long <- mutate(long, cell_id = rownames(long))}
  
  return(long)
}

# make everything long
crab_long <- make.long(crab, "not", "count")
points_long <- make.long(points, "flat", "")
depth_long <- make.long(depth, "flat", "")
winter_mean_long <- make.long(winter_mean, "not", "winter_mean")
summer_mean_long <- make.long(summer_mean, "not", "summer_mean")
summer_max_long <- make.long(summer_max, "not", "summer_max")
winter_min_long <- make.long(winter_min, "not", "winter_min")
yearly_flags <- make.long(yearly_flags, "not", "yearly_flags")
ddd <- make.long(ddd, "not", "ddd")
rdd <- make.long(rdd, "not", "rdd")
sdd <- make.long(sdd, "not", "sdd")
fall_flags <- make.long(fall_flags, "not", "fall_flags")
spring_flags <- make.long(spring_flags, "not", "spring_flags")
winter_flags <- make.long(winter_flags, "not", "winter_flags")

pop_long <- make.long(pop, "flat", "")

# join data
data <- crab_long %>%
  left_join(points_long, by = "cell_id") %>%
  left_join(depth_long, by = "cell_id") %>% rename("depth" = "wide") %>%
  left_join(winter_mean_long, by = c("cell_id", "year")) %>%
  left_join(summer_mean_long, by = c("cell_id", "year")) %>%
  left_join(summer_max_long, by = c("cell_id", "year")) %>%
  left_join(winter_min_long, by = c("cell_id", "year")) %>%
  left_join(rdd, by = c("cell_id", "year")) %>%
  left_join(sdd, by = c("cell_id", "year")) %>%
  left_join(ddd, by = c("cell_id", "year")) %>%
  left_join(fall_flags, by = c("cell_id", "year")) %>%
  left_join(spring_flags, by = c("cell_id", "year")) %>%
  left_join(winter_flags, by = c("cell_id", "year")) %>%
  left_join(yearly_flags, by = c("cell_id", "year")) %>%
  left_join(pop_long, by = "cell_id") %>% rename("pop" = "wide")
```

```{r}
#| label: function to extract model selection stats from a given model

glmmtmb_stats <- function(model, mod_string) {
  stats <- data.frame("Model" = mod_string, 
                    "Intercept" = as.numeric(fixef(model)$cond[1]),
                    "Pearson" = cor(model.frame(model)[[1]],fitted(model),method="pearson"),
                    "Spearman" = cor(model.frame(model)[[1]],fitted(model),method="spearman"),
                    "RMSE" = sqrt(mean(residuals (model, type = "response")^2)),
                    "MAE" = mean(abs(residuals(model, type = "response"))),
                    "AIC" = AIC(model), "BIC" = BIC(model),
                    "log_link" = family(model)$link, "df" = attr(logLik(model), "df"),
                    "Package" = "glmmTMB")
  return(stats)}
```

## Null model:

```{r}
#| label: null model

null <- glmmTMB(count ~ 1, ziformula = ~ pop + latitude, family = nbinom2(), data = data)
null_stats <- glmmtmb_stats(null, "Null")
```

## One covariate at a time:

```{r}
#| label: one covariate at a time, unscaled

# year <- glmmTMB(count ~ year, ziformula = ~ pop + latitude, family = nbinom2(), data = data) # convervence issues, reject

depth.mod <- glmmTMB(count ~ depth, ziformula = ~ pop + latitude, family = nbinom2(), data = data)
plot(simulateResiduals(depth.mod)) # okay
depth_stats <- glmmtmb_stats(depth.mod, "Depth")

winter_mean.mod <- glmmTMB(count ~ winter_mean, ziformula = ~ pop + latitude, family = nbinom2(), data = data)
plot(simulateResiduals(winter_mean.mod)) # okay, but did hit an error for missing values
winter_mean_stats <- glmmtmb_stats(winter_mean.mod, "Winter Mean")

summer_mean.mod <- glmmTMB(count ~ summer_mean, ziformula = ~ pop + latitude, family = nbinom2(), data = data)
plot(simulateResiduals(summer_mean.mod)) # quantile deviations, don't reject but probably bad
summer_mean_stats <- glmmtmb_stats(summer_mean.mod, "Summer Mean")

summer_max.mod <- glmmTMB(count ~ summer_max, ziformula = ~ pop + latitude, family = nbinom2(), data = data)
plot(simulateResiduals(summer_max.mod)) # okay
summer_max_stats <- glmmtmb_stats(summer_max.mod, "Summer Max")

winter_min.mod <- glmmTMB(count ~ winter_min, ziformula = ~ pop + latitude, family = nbinom2(), data = data)
plot(simulateResiduals(winter_min.mod)) # okay
winter_min_stats <- glmmtmb_stats(winter_min.mod, "Winter Min")

# rdd.mod <- glmmTMB(count ~ rdd, ziformula = ~ pop + latitude, family = nbinom2(), data = data)
# plot(simulateResiduals(rdd.mod)) # failed one test that i can't see and threw warning may have inflated type I errors, reject

# sdd.mod <- glmmTMB(count ~ sdd, ziformula = ~ pop + latitude, family = nbinom2(), data = data) # dropped the column, reject

ddd.mod <- glmmTMB(count ~ ddd, ziformula = ~ pop + latitude, family = nbinom2(), data = data)
plot(simulateResiduals(ddd.mod)) # okay
ddd_stats <- glmmtmb_stats(ddd.mod, "DDD")

# fall_flags.mod <- glmmTMB(count ~ fall_flags, ziformula = ~ pop + latitude, family = nbinom2(), data = data) # dropped the column, reject

spring_flags.mod <- glmmTMB(count ~ spring_flags, ziformula = ~ pop + latitude, family = nbinom2(), data = data)
plot(simulateResiduals(spring_flags.mod)) # okay
spring_flags_stats <- glmmtmb_stats(spring_flags.mod, "Spring Flags")

winter_flags.mod <- glmmTMB(count ~ winter_flags, ziformula = ~ pop + latitude, family = nbinom2(), data = data)
plot(simulateResiduals(winter_flags.mod)) # okay
winter_flags_stats <- glmmtmb_stats(winter_flags.mod, "Winter Flags")

yearly_flags.mod <- glmmTMB(count ~ yearly_flags, ziformula = ~ pop + latitude, family = nbinom2(), data = data)
plot(simulateResiduals(yearly_flags.mod)) # okay
yearly_flags_stats <- glmmtmb_stats(yearly_flags.mod, "Yearly Flags")
```

```{r}
#| label: combine single-covariate models, unscaled

single_covariates <- rbind(null_stats, depth_stats, summer_mean_stats, winter_mean_stats, summer_max_stats, winter_min_stats, ddd_stats, winter_flags_stats, spring_flags_stats, yearly_flags_stats) # not sure what to make of this - summer_mean model is much better but there are quantile deviations in the qq plot
```

```{r}
#| label: one covariate at a time, scaled

depth.mods <- glmmTMB(count ~ scale(depth), ziformula = ~ scale(pop) + scale(latitude), family = nbinom2(), data = data)
plot(simulateResiduals(depth.mods)) # okay
depth_stats <- glmmtmb_stats(depth.mods, "Depth")

winter_mean.mods <- glmmTMB(count ~ scale(winter_mean), ziformula = ~ scale(pop) + scale(latitude), family = nbinom2(), data = data)
plot(simulateResiduals(winter_mean.mods)) # okay, but did hit an error for missing values
winter_mean_stats <- glmmtmb_stats(winter_mean.mods, "Winter Mean")

summer_mean.mods <- glmmTMB(count ~ scale(summer_mean), ziformula = ~ scale(pop) + scale(latitude), family = nbinom2(), data = data)
plot(simulateResiduals(summer_mean.mods)) # quantile deviations, don't reject but probably bad
summer_mean_stats <- glmmtmb_stats(summer_mean.mods, "Summer Mean")

summer_max.mods <- glmmTMB(count ~ scale(summer_max), ziformula = ~ scale(pop) + scale(latitude), family = nbinom2(), data = data)
plot(simulateResiduals(summer_max.mods)) # okay
summer_max_stats <- glmmtmb_stats(summer_max.mods, "Summer Max")

winter_min.mods <- glmmTMB(count ~ scale(winter_min), ziformula = ~ scale(pop) + scale(latitude), family = nbinom2(), data = data)
plot(simulateResiduals(winter_min.mods)) # okay
winter_min_stats <- glmmtmb_stats(winter_min.mods, "Winter Min")

ddd.mods <- glmmTMB(count ~ scale(ddd), ziformula = ~ scale(pop) + scale(latitude), family = nbinom2(), data = data)
plot(simulateResiduals(ddd.mods)) # okay
ddd_stats <- glmmtmb_stats(ddd.mods, "DDD")

spring_flags.mods <- glmmTMB(count ~ scale(spring_flags), ziformula = ~ scale(pop) + scale(latitude), family = nbinom2(), data = data)
plot(simulateResiduals(spring_flags.mods)) # okay
spring_flags_stats <- glmmtmb_stats(spring_flags.mods, "Spring Flags")

winter_flags.mods <- glmmTMB(count ~ scale(winter_flags), ziformula = ~ scale(pop) + scale(latitude), family = nbinom2(), data = data)
plot(simulateResiduals(winter_flags.mods)) # okay
winter_flags_stats <- glmmtmb_stats(winter_flags.mods, "Winter Flags")

yearly_flags.mods <- glmmTMB(count ~ scale(yearly_flags), ziformula = ~ scale(pop) + scale(latitude), family = nbinom2(), data = data)
plot(simulateResiduals(yearly_flags.mods)) # okay
yearly_flags_stats <- glmmtmb_stats(yearly_flags.mods, "Yearly Flags")
```

```{r}
#| label: combine single-covariate models, scaled

scaled_single_covariates <- rbind(null_stats, depth_stats, summer_mean_stats, winter_mean_stats, summer_max_stats, winter_min_stats, ddd_stats, winter_flags_stats, spring_flags_stats, yearly_flags_stats) # not sure what to make of this - summer_mean model is much better but there are quantile deviations in the qq plot
```

## Response curves for single-covariate models:

```{r}
#| label: response and response curves functions

response <- function(model, covar_vect, covar_name) { # model expects model object, covar_vect expects a column of the og dataframe, covar_name expects a the string of the variable name used to fit the model
  
  ## prepare
  covar_seq <- seq(from=min(scale(covar_vect)[,1]), to=max(scale(covar_vect)[,1]), length.out=100) # sequence of values for the covariate you're looking for the response to
  pred_dat <- cbind(as.data.frame(covar_seq),
                    mean(scale(data$latitude)), 
                    mean(scale(data$pop))) # combine with means of other necessary params
  colnames(pred_dat) <- c(covar_name, "latitude", "pop") # assign column names
  
  ## predict
  preds <- predict(model, newdata = pred_dat, se.fit = T, type = "link") # predict abuns
  preds_df <- cbind(as.numeric(preds$fit), as.numeric(preds$se.fit), pred_dat) # make df
  colnames(preds_df) <- c("pred_link", "se_link", "covariate", "latitude", "pop") # set colnames again
  preds_df <- preds_df %>% 
    mutate(pred_resp = exp(pred_link),
           lower_resp = exp(pred_link - 1.96 * se_link), 
           upper_resp = exp(pred_link + 1.96 * se_link)) %>% # find upper/lower
    mutate(covar_descaled = covar_seq * attr(scale(covar_vect), 'scaled:scale') +
                               attr(scale(covar_vect), 'scaled:center'))
  
  ## return
  return(preds_df)
}

response_curve <- function(preds_df, xlab, title) {
  plot <- ggplot(preds_df, aes(x=covariate, y=pred_resp)) + # covariate will be on the link scale, pred_resp will be back on the response scale
    geom_line() + # mu line
    geom_ribbon(aes(ymin=lower_resp, ymax=upper_resp), alpha=0.3) + # 95% CI on response scale
    scale_x_continuous(breaks = preds_df$covariate[nice_ids], # breaks from scaled covar
    labels = round(preds_df$covar_descaled[nice_ids])) + # labels from descaled covar
    labs(y="Crab abundance", x=xlab, title=title) +
  theme_classic()
  return(plot)
}

# example
# response_curve(response(winter_min.mods, data$winter_min, "winter_min"), "Minimum winter temperature (°C)", "Response to minimum winter temperature")
```

```{r}
#| label: curves for scaled, single -covariate models

winter_min_plot <- response_curve(response(winter_min.mods, data$winter_min, "winter_min"), "Minimum winter temperature (°C)", "7. Minimum winter temperature")
winter_mean_plot <- response_curve(response(winter_mean.mods, data$winter_mean, "winter_mean"), "Mean winter temperature (°C)", "6. Mean winter temperature")
summer_mean_plot <- response_curve(response(summer_mean.mods, data$summer_mean, "summer_mean"), "Mean summer temperature (°C)", "1. Mean summer temperature")
summer_max_plot <- response_curve(response(summer_max.mods, data$summer_max, "summer_max"), "Maximum summer temperature (°C)", "2. Maximum summer temperature")
winter_min_plot <- response_curve(response(winter_min.mods, data$winter_min, "winter_min"), "Minimum winter temperature (°C)", "Minimum winter temperature")
depth_plot <- response_curve(response(depth.mods, data$depth, "depth"), "Depth (m)", "3. Depth")
ddd_plot <- response_curve(response(ddd.mods, data$ddd, "ddd"), "Yearly death degree days (3°C and below)", "8. Death degree days")
winter_flags_plot <- response_curve(response(winter_flags.mods, data$winter_flags, "winter_flags"), "Days below 3°C during winter", "5. Winter flag days")
spring_flags_plot <- response_curve(response(spring_flags.mods, data$spring_flags, "spring_flags"), "Days below 3°C during spring", "4. Spring flag days")
yearly_flags_plot <- response_curve(response(yearly_flags.mods, data$yearly_flags, "yearly_flags"), "Days below 3°C for that year starting in December", "9. Yearly flag days")

(summer_mean_plot + summer_max_plot + depth_plot) / (spring_flags_plot + winter_flags_plot + winter_mean_plot) / (winter_min_plot + ddd_plot + yearly_flags_plot) + plot_annotation(title = "Single-covariate models", subtitle = "Ranked by AIC")
```

## Adding random effects

Best single-covariate (fixed effect or "best optimal" termed by Zuur) model is:

optimal \<- glmmTMB(count \~ scale(summer_mean), ziformula = \~ scale(pop) + scale(latitude), family = nbinom2(), data = data)

```{r}
#| label: build from best fixed effect model

optimal <- glmmTMB(count ~ scale(summer_mean), ziformula = ~ scale(pop) + scale(latitude), family = nbinom2(), data = data)
plot(simulateResiduals(optimal)) # quantile deviations
optimal_stats <- glmmtmb_stats(optimal, "Optimal with ZI formula")

cellfx <- glmmTMB(count ~ scale(summer_mean) + 1|cell_id, ziformula = ~ scale(pop) + scale(latitude), family = nbinom2(), data = data)
plot(simulateResiduals(cellfx)) # failed dispersion test
cellfx_stats <- glmmtmb_stats(cellfx, "Cell random intercepts")

yrfx <- glmmTMB(count ~ scale(summer_mean) + (1|year), ziformula = ~ scale(pop) + scale(latitude), family = nbinom2(), data = data)
plot(simulateResiduals(yrfx)) # no problems
yrfx_stats <- glmmtmb_stats(yrfx, "Year random intercepts")

fx_stats <- rbind(optimal_stats, cellfx_stats, yrfx_stats) # year random effects are best. this is great! makes sense.

# yrcellfx <- glmmTMB(count ~ scale(summer_mean) + year|cell_id, ziformula = ~ scale(pop) + scale(latitude), family = nbinom2(), data = data) # let it run 8 minutes before stopping it, at which point i got NA/NaN function evaluation errors, reject

## testing combinations of removing the zi formula, all these models fail the qq plot dispersion test

## with an offset but without zi formula, with the best random effect found above

# yrfx_off <- glmmTMB(count ~ scale(summer_mean) + (1|year) + offset(log(pop)), ziformula = ~ scale(latitude), family = nbinom2(), data = data)
# plot(simulateResiduals(yrfx_off)) # failed dispersion test
# yrfx_off_stats <- glmmtmb_stats(yrfx_off, "Year random intercepts with population offset")
# 
# yrfx_off_nozi <- glmmTMB(count ~ scale(summer_mean) + (1|year) + offset(log(pop)), ziformula = ~ 1, family = nbinom2(), data = data)
# plot(simulateResiduals(yrfx_off_nozi)) # failed dispersion test
# yrfx_off_nozi_stats <- glmmtmb_stats(yrfx_off_nozi, "Year random intercepts with population offset and without zero-inflation formula")

## without zi formula, just to check if that was subbing in for random effects somehow

# optimal_sanszi <- glmmTMB(count ~ scale(summer_mean), ziformula = ~ 1, family = nbinom2(), data = data)
# plot(simulateResiduals(optimal_sanszi)) # more quantile deviations
# optimal_stats <- glmmtmb_stats(optimal_sanszi, "Optimal without ZI formula")
# 
# popfx_sanszi <- glmmTMB(count ~ scale(summer_mean) + 1|pop, ziformula = ~ 1, family = nbinom2(), data = data)
# plot(simulateResiduals(popfx_sanszi)) # failed dispersion test
# popfx_stats <- glmmtmb_stats(popfx_sanszi, "Population effect without ZI formula")
# 
# latfx_sanszi <- glmmTMB(count ~ scale(summer_mean) + 1|latitude, ziformula = ~ 1, family = nbinom2(), data = data)
# plot(simulateResiduals(latfx_sanszi)) # failed dispersion test
# latfx_stats <- glmmtmb_stats(latfx_sanszi, "Latitude effect without ZI formula")
# 
# ## with only half of the zi formula, similar sanity check to the above
# 
# popfx <- glmmTMB(count ~ scale(summer_mean) + 1|pop, ziformula = ~ scale(latitude), family = nbinom2(), data = data)
# plot(simulateResiduals(popfx)) # failed dispersion test
# popfx_withzi_stats <- glmmtmb_stats(popfx, "Population effect with latitude ZI formula")
# 
# latfx <- glmmTMB(count ~ scale(summer_mean) + 1|latitude, ziformula = ~ scale(pop), family = nbinom2(), data = data)
# plot(simulateResiduals(latfx)) # failed dispersion test
# latfx_withzi_stats <- glmmtmb_stats(latfx, "Latitude effect with population ZI formula")
```

## Add additional covariates

Best random-effect model with a single fixed effect is:

effect \<- glmmTMB(count \~ scale(summer_mean) + (1\|year), ziformula = \~ scale(pop) + scale(latitude), family = nbinom2(), data = data)

```{r}
#| label: adding covariates

# return to scaled_single_covariates stats dataframe. we only want to add covariates that aren't collinear and that beat the null model. the single-covariate models that beat the null were summer mean, summer max, and depth

# test if summer max is collinear with the covariate we already have (summer mean)
cor(data$summer_max, data$summer_mean) # 0.94 means they are HIHGLY correlated, won't try adding summer max

# so we just have to test if depth helps things out!
effect <- glmmTMB(count ~ scale(summer_mean) + (1|year), ziformula = ~ scale(pop) + scale(latitude), family = nbinom2(), data = data)
plot(simulateResiduals(effect)) # looks okay
effect_stats <- glmmtmb_stats(effect, "Optimal model with random effect")

effect_wdepth <- glmmTMB(count ~ scale(summer_mean) + scale(depth) + (1|year), ziformula = ~ scale(pop) + scale(latitude), family = nbinom2(), data = data)
plot(simulateResiduals(effect_wdepth)) # looks okay
effect_wdepth_stats <- glmmtmb_stats(effect_wdepth, "Optimal model with random effect and additional depth covariate")

View(rbind(effect_stats, effect_wdepth_stats)) # BIC gets worse and AIC only gets one unit better, which isn't enough to justify adding the covariate
```

[Note to self:]{.underline} AIC of ideal model is 1302
